Forecasting interrupted time series

### 2024-01-08 fits.qmd
Added discussion

>>>>

# Discussion

We have presented several approaches to forecasting in the presence of an interruption. Each has its advantages and disadvantages, and the choice of which to use will depend on the situation. In fact, we have used almost all of these approaches in our own consulting work. An intervention model is often a good solution, provided the intervention can be modelled relatively simply. However, if the intervention is complex, then a highly adaptive method is often better. The highly adaptive model is also useful when the timing of the interruption is unknown, or when the post-interruption period is expected to be very different from the pre-interruption period. The missing value approach is particularly useful when only post-interruption forecasts are required, and forecasts during the interruption period are not needed.

We have not previously used the weighted approach in practice, and software to implement it is not currently available. However, it seems to be a good compromise between the highly adaptive model and the intervention model, and we expect it to be useful in many situations, once software is available to implement it easily.

Finally, the ensemble approach is useful when there is uncertainty about which approach to use, and will often result in greater accuracy due the power of averaging.

We have not discussed probabilistic forecasting in any detail in this paper, other than producing some prediction intervals for some of the examples. One would expect forecast uncertainty to increase during, and to some extent after, a major disruption. The highly adaptive model and innovation model should produce relatively good prediction intervals, although the variance is often underestimated. Because the missing value approach is producing counterfactual ("what might have been") forecasts, the prediction intervals will often not include the actual observations during the disruption period. Prediction intervals for ensemble forecasts can be obtained by averaging the end points of the component prediction intervals [@lichtendahl2013better], for those models where the prediction intervals seem reasonable.

### 2024-01-08 fits.tex
Added discussion

@@

\subsection{Use a dynamic regression model with intervention
covariates}\label{use-a-dynamic-regression-model-with-intervention-covariates}

@@

This has the advantage of retaining the memory of the past, and so the
seasonal patterns and other dynamics will be retained. However, the
model will assume that the post-interruption period will be similar to
the pre-interruption period, and so the prediction intervals may be be
too narrow.

Advantages: retains full memory of the past, and allows the change
period to be effectively modelled provided you choose the covariates
well.

Disadvantages: requires a lot of thought to choose the covariates well.
Assumes that the post-pandemic period will be similar to the
pre-pandemic period.

\subsection{Treat the covid period as missing and use a model that
handles missing
values}\label{treat-the-covid-period-as-missing-and-use-a-model-that-handles-missing-values}

@@

\subsection{Estimate what might have been and adjust the
data}\label{estimate-what-might-have-been-and-adjust-the-data}

A fourth solution is to estimate what might have been during the period
of disruption, and then use the adjusted data to fit a model. The
estimates could be made using any convenient method. One approach would
be to set the estimates to equal the forecasts made using only
pre-interruption data. This then becomes almost equal to the previous
solution except that the prediction intervals will be narrower because
the estimation uncertainty has not been taken into account.

\subsection{Over sampling (weighting)}\label{over-sampling-weighting}

Oversampling is a technique used in machine learning to address class
imbalance within a dataset. In this context, the goal is to ensure that
the model is more influenced by the patterns observed before the
disruption period while still accounting for the observations during the
disruption period.

Using this weighting approach, we can leverage the insights from the
pre-disruptive period while still accounting for the disruptive period's
influence. It allows the model to better capture the underlying patterns
of the time series data, considering both periods, and potentially
producing more accurate forecasts.

Advantages:

Preserves pre-event patterns: By assigning higher weights to the
pre-disruption time series data, the model is encouraged to focus more
on the patterns observed before the disruptive event. This can be
beneficial in capturing the underlying dynamics of the time series that
might be essential for accurate forecasting in the post-disruption era.

Robustness to disruption impact: The weighting approach allows the model
to still take the disruption period into account, even if it's
considered less influential. This ensures that the model is not
completely blind to the disruption's effects and can adapt to changes or
shifts in the time series caused by the event.

Easy to Implement: ??

Interpretability?:

Disadvantages:

@@

\section{Examples}\label{examples}

Let's consider some examples

\subsection{Australian tourism}\label{australian-tourism}

@@

We apply the four solutions discussed in the previous section to these
data, making 12 month forecasts at the end of each year from 2019 to
2022. We fit ETS, ARIMA and dynamic ARIMA models to the data
\autocite{fpp3}, first applying a log transformation to ensure the
resulting forecasts are positive. This is possible, because the
observations never reach exactly zero, with the smallest number of
visitors per month equal to 2250 in April 2020. For all forecasts, we
also show 90\% prediction intervals.

@@

@@

\caption{\label{fig-tsol2-plot}Solution 2. We use a left shift from
March 2020 to October 2022, and a ramp from October 2021 to October
2022. After October 2022, there are no non-zero predictors.}

@@

The third solution involved setting the observations during the
disruption period to missing, and then fitting an ARIMA model to the
series. Consequently, the first three years of forecasts show what might
have been without the COVID-19 pandemic, based on the history to the end
of 2019. The final forecasts for 2023 use the data to the end of 2019,
and the two observations in November and December 2022. These are much
better, but the prediction intervals are too narrow, as there hasn't
been sufficient recent data.

@@

Finally, we show in Figure~\ref{fig-tsol4-plot} the forecasts obtained
using solution 4. Here we have replaced observations between March 2020
and October 2022 with estimates based on the average of the same month
in the three years prior to March 2020. The resulting forecasts are
similar to those from solution 3, but with narrower prediction intervals
because the model is (falsely) assuming that the ``observations'' during
the disruption are real.

@@

We have also shown the periods of lockdown due to COVID-19. In
Melbourne, there were six separate lockdowns ranging from 5 days to 111
days in length. The first official lockdown period was from 31 March
2020 to 12 May 2020, and the last lockdown was from 05 Aug 2021 to 21
Oct 2021. However, the first period was preceded by over a week when
most people elected to stay and work at home, so we have chosen to start
the first period on 23 Mar 2020, to better reflect human behaviour.
Otherwise, we have used the official lockdown dates.

We apply the four solutions discussed in Section~\ref{sec-methods}. We
fit ETS, ARIMA and dynamic ARIMA models to the data, with no
transformations used. We have evaluated the methods by using time series
cross-validation, with the initial training set comprising the whole of
2019, and subsequent training sets growing by 1 week at a time, to the
end of 2021. The test sets are always 3 weeks long. Thus, we have
evaluated the forecasts during 2020 and 2021, covering all lockdown
periods, and the start of the recovery period. The results are shown in
\textbf{?@fig-walkers-plot1}.

\subsection{Example: Ambulance
attendances}\label{example-ambulance-attendances}

\includegraphics{fits_files/figure-pdf/plot-data-1.pdf}

\includegraphics{fits_files/figure-pdf/plot-data-2.pdf}

\includegraphics{fits_files/figure-pdf/unnamed-chunk-10-1.pdf}

\includegraphics{fits_files/figure-pdf/forecast-1.pdf}

\includegraphics{fits_files/figure-pdf/forecast-2.pdf}

\includegraphics{fits_files/figure-pdf/forecast-3.pdf}

>>>>

The disruption may result in relatively simple changes in the series;
for example a level shift at the start of the disruption, and another at
the end of the disruption. Or they may be more complex, with changes to
the seasonal patterns, and changes to the level, which evolve over time.
In this paper, we consider a range of models that can be used to handle
such changes, and compare their performance on some real data sets.

We introduce several approaches to handling interruptions when
forecasting in Section~\ref{sec-methods}. We then apply these approaches
to two real data sets in \textbf{?@sec-sec-examples}. Finally, in
\textbf{?@sec-conclusions}, we discuss some of the advantages and
disadvantages of each approach, and provide some recommendations for
practitioners.

@@

\subsection{Use an intervention model}\label{use-an-intervention-model}

@@

This has the advantage of retaining the memory of the past, and so the
seasonal patterns and other dynamics will be retained. This allows the
change period to be effectively modelled provided the intervention
variables are chosen well. However, the model will assume that the
post-interruption period will be similar to the pre-interruption period,
and so the prediction intervals may be be too narrow, especially if
there is a lasting effect beyond the end of the interruption.

\subsection{Treat the interruption period as
missing}\label{treat-the-interruption-period-as-missing}

@@

\subsection{Estimate what might have
been}\label{estimate-what-might-have-been}

A fourth solution is to estimate what might have been during the period
of disruption, and then use the adjusted data to fit a model. The
estimates could be made using any convenient method. One approach would
be to set the estimates to equal the forecasts made using only
pre-interruption data. This then becomes almost equal to the previous
solution except that the prediction intervals will be narrower because
the estimation uncertainty has not been taken into account. On the other
hand, it is more flexible than the previous approach because models that
do not handle missing values can then be used.

An alternative would be to take the model estimated under the previous
solution (setting the observations during the disruption to missing),
and use it to estimate what might have been during the disruption. Then
the model can be re-estimated using the adjusted data. Note that
forecasts obtained in this way during the disruption will not be true
forecasts, because they will have used data from the future in computing
the adjusted data during the disruption. But post-disruption forecasts
will be true forecasts, and should be almost the same as those obtained
using the previous solution.

\subsection{Downweight the interruption
period}\label{downweight-the-interruption-period}

Intuitively, we want the model to be more influenced by the patterns
observed before the disruption period than those during the disruption,
but we still want to take some account of the observations during the
disruption period.

Using this weighting approach, we can leverage the insights from the
pre-disruptive period while still accounting for the disruptive period's
influence. It allows the model to better capture the underlying dynamics
of the time series in non-disrupted periods, which may be essential for
accurate forecasting in the post-disruption era.

If the weights during the disruption are set to zero, and the weights at
other times are set to 1, this becomes equivalent to the missing value
solution above. Rather than ignoring the data during the disruption, the
weighting approach allows the model to not be completely blind to the
disruption's effects, and adapt to changes or shifts in the time series
caused by the event.

While conceptually simple to implement, in practice, this approach may
require more work than the other methods discussed here as most
forecasting software does not allow for the explicit use of weights.

@@

As with may forecasting problems, taken an ensemble approach often leads
to more accurate forecasts \autocite{combinations}. Here, we could
combine some or all four of the approaches discussed above to obtain an
ensemble forecast. In fact, if we were unable to implement the weighted
approach, by using an ensemble of the other approaches we are
effectively downweighting the observations during the disruption period
as they are only explicitly used in the first two approaches (using a
highly adaptive model, or using an intervention model).

However, a disadvantage of this approach is that it is averaging
forecasts that aim to achieve different objectives. For example, the
highly adaptive model and the intervention model are trying to forecast
what happened during the disruption, while the missing value and
estimation approaches are trying to forecast what might have happened.
So the ensemble approach will lie somewhere between these two
objectives.

\section{Examples}\label{sec-examples}

Let's consider some examples to data observed during the last few years,
where the effect of the COVID-19 pandemic has been particularly evident.

\subsection{Australian tourism}\label{sec-tourism}

@@

We apply the first four solutions discussed in the previous section to
these data, making 12 month forecasts at the end of each year from 2019
to 2022. We fit ETS, ARIMA and dynamic ARIMA models to the data
\autocite{fpp3}, first applying a log transformation to ensure the
resulting forecasts are positive. This is possible, because the
observations never reach exactly zero, with the smallest number of
visitors per month equal to 2250 in April 2020. For all forecasts, we
also show 90\% prediction intervals.

@@

\begin{figure}[b]

@@

\caption{\label{fig-tsol2-plot}Solution 2. We use a level shift from
March 2020 to October 2022, and a ramp from October 2021 to October
2022. After October 2022, all intervention variables are set to zero.}

@@

The third solution (Figure~\ref{fig-tsol3-plot}) involved setting the
observations during the disruption period to missing, and then fitting
an ARIMA model to the series. Consequently, the first three years of
forecasts show what might have been without the COVID-19 pandemic, based
on the history to the end of 2019. The final forecasts for 2023 use the
data to the end of 2019, and the two observations in November and
December 2022. These are much better, but the prediction intervals are
too narrow, as there hasn't been sufficient recent data.

\begin{figure}[!b]

@@

We show in Figure~\ref{fig-tsol4-plot} the forecasts obtained using
solution 4. Here we have replaced observations between March 2020 and
October 2022 with estimates based on the average of the same month in
the three years prior to March 2020. The resulting forecasts are similar
to those from solution 3, but with narrower prediction intervals because
the model is (falsely) assuming that the ``observations'' during the
disruption are real.

Finally, in Figure~\ref{fig-tensemble-plot}, we show the ensemble
forecasts obtained by averaging the forecasts from the four solutions
discussed above. The first and fourth solutions carry double weight,
because both the ETS and ARIMA forecasts were included in the ensemble.
This time, no prediction intervals have been computed. The first set of
forecasts have not taken any account of the disruption, and so are
almost identical to those obtained in the previous plots. The last set
of forecasts are reasonably good, showing what is obtained by combining
the forecasting methods used to handle the disruption period. However,
the forecasts for 2021 and 2022 are particularly poor, because they are
averaging forecasts that aim to achieve different objectives, and are
affected by the poor forecasts obtained using the dynamic regression
model.

\includegraphics{fits_files/figure-pdf/fig-tensemble-plot-1.pdf}

\caption{\label{fig-tensemble-plot}Ensemble solution, combining the
previous four approaches shown in Figures
\ref{fig-tsol1-plot}--\ref{fig-tsol4-plot}.}

@@

We have also shown the periods of lockdown due to COVID-19. In
Melbourne, there were six separate lockdowns ranging from 5 days to 111
days in length. The first official lockdown period was from 31 March
2020 to 12 May 2020, and the last lockdown was from 05 Aug 2021 to 21
Oct 2021. However, the first period was preceded by over a week when
most people elected to stay and work at home, so we have chosen to start
the first period on 23 Mar 2020, to better reflect human behaviour.
Otherwise, we have used the official lockdown dates
\autocite{MelbourneLockdowns}.

We evaluate the solutions discussed in Section~\ref{sec-methods} by
using time series cross-validation, with the initial training set
comprising the whole of 2019, and subsequent training sets growing by 1
week at a time, to the end of 2021. The test sets are always 1 week
long. Thus, we have evaluated the forecasts during 2020 and 2021,
covering all lockdown periods, and the start of the recovery period. No
transformations of the data have been used. The results are shown in
Figures \ref{fig-walkers-plot1}--\ref{fig-walkers-plot4}. Prediction
intervals have not been shown to avoid cluttering the plots.

\includegraphics{fits_files/figure-pdf/fig-walkers-plot1-1.pdf}

\caption{\label{fig-walkers-plot1}Solution 1: the forecasts do not take
into account the lockdown periods, but the ARIMA model is largely able
to adjust to the changing level of the series.}

We use ARIMA models in Figure~\ref{fig-walkers-plot1}, which have mostly
been able to adapt to the changing level of the series, apart from
during the first few weeks of the first lockdown, and the first day or
two of subsequent lockdowns.

In Figure~\ref{fig-walkers-plot2}, we have used an intervention model,
with a dummy variable indicating the lockdown periods. This has worked a
little better than the ARIMA model used in
Figure~\ref{fig-walkers-plot1}, apart from around the first lockdown
period. Many people were in self-imposed lockdown before the start of
the first lockdown period, and were reluctant to return to the city
after it ended, resulting in poor forecasts on either side of the first
lockdown.

\includegraphics{fits_files/figure-pdf/fig-walkers-plot2-1.pdf}

\caption{\label{fig-walkers-plot2}Solution 2: the forecasts are based on
an ARIMA model with a dummy variable indicating when the lockdown
periods occurred. Many people were in self-imposed lockdown before the
start of the first lockdown period, and were reluctant to return to the
city after it ended, resulting in poor forecasts on either side of the
first lockdown. The model was able to adjust to the changing level of
the series during subsequent lockdowns.}

\includegraphics{fits_files/figure-pdf/fig-walkers-plot3-1.pdf}

\caption{\label{fig-walkers-plot3}Solution 3: the lockdown periods are
set to missing, and the ARIMA model uses the remaining data to produce
forecasts. Thus, the forecasts are based on only previous non-lockdown
data. Naturally, the forecasts during the lockdown periods do not
reflect reality, but they can be interpreted as what might have been.}

\includegraphics{fits_files/figure-pdf/fig-walkers-plot4-1.pdf}

\caption{\label{fig-walkers-plot4}Solution 4: the lockdown periods are
replaced with estimated counts based on an ARIMA model applied to the
remaining data. Then a new ARIMA model is fitted to the whole data set,
and used to produce forecasts. As a result, the forecasts during the
lockdown periods are not true forecasts, as they use non-lockdown data
from the future.}

\includegraphics{fits_files/figure-pdf/fig-walkers-ensemble-plot-1.pdf}

\caption{\label{fig-walkers-ensemble-plot}Ensemble forecasts based on
the first two solutions shown in Figures \ref{fig-walkers-plot1} and
\ref{fig-walkers-plot2}. Other than having some difficulty around the
period of the first lockdown, the results are relatively good.}

Figure~\ref{fig-walkers-plot3} shows the results obtained by setting the
observations during the lockdown periods to missing, and then fitting an
ARIMA model to the remaining data. The forecasts during lockdowns show
what might have been if that particular lockdown hadn't occurred, based
on previous non-lockdown data.

The forecasts shown in Figure~\ref{fig-walkers-plot4} are obtained by
replacing the observations during the lockdown periods with estimates
based on an ARIMA model applied to the remaining data. Then a new ARIMA
model is fitted to the whole data set, and used to produce forecasts. As
a result, the ``forecasts'' appear to interpolate across the lockdown
periods, reflecting neither the true lockdown pattern, or the pattern
that might have occurred without the lockdowns.

Finally, in Figure~\ref{fig-walkers-ensemble-plot}, we show the ensemble
forecasts obtained by averaging the forecasts from the first two
solutions discussed above. The results are relatively good, other than
having some difficulty around the period of the first lockdown.

\section{Discussion}\label{discussion}

We have presented several approaches to forecasting in the presence of
an interruption. Each has its advantages and disadvantages, and the
choice of which to use will depend on the situation. In fact, we have
used almost all of these approaches in our own consulting work. An
intervention model is often a good solution, provided the intervention
can be modelled relatively simply. However, if the intervention is
complex, then a highly adaptive method is often better. The highly
adaptive model is also useful when the timing of the interruption is
unknown, or when the post-interruption period is expected to be very
different from the pre-interruption period. The missing value approach
is particularly useful when only post-interruption forecasts are
required, and forecasts during the interruption period are not needed.

We have not previously used the weighted approach in practice, and
software to implement it is not currently available. However, it seems
to be a good compromise between the highly adaptive model and the
intervention model, and we expect it to be useful in many situations,
once software is available to implement it easily.

Finally, the ensemble approach is useful when there is uncertainty about
which approach to use, and will often result in greater accuracy due the
power of averaging.

We have not discussed probabilistic forecasting in any detail in this
paper, other than producing some prediction intervals for some of the
examples. One would expect forecast uncertainty to increase during, and
to some extent after, a major disruption. The highly adaptive model and
innovation model should produce relatively good prediction intervals,
although the variance is often underestimated. Because the missing value
approach is producing counterfactual (``what might have been'')
forecasts, the prediction intervals will often not include the actual
observations during the disruption period. Prediction intervals for
ensemble forecasts can be obtained by averaging the end points of the
component prediction intervals, for those models where the prediction
intervals seem reasonable.
